---
title: "A SEM user's guide to dagitty for R"
author: "Johannes Textor"
date: "`r Sys.Date()`"
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{A SEM user's guide to dagitty for R}
  \usepackage[utf8]{inputenc} 
output:
  knitr:::html_vignette:
    toc: yes
---

```{r, echo = FALSE, message = FALSE}
knitr::opts_chunk$set(comment = "")
library(dagitty)
```

## What is dagitty

Dagitty is a software to analyze causal diagrams, also known as directed acyclic graphs 
(DAGs). Structural equation models (SEMs) can be viewed as a parametric form of DAGs,
which encode linear functions instead of arbitrary nonlinear functions.

Because every SEM is a DAG, much of the methodology developed for DAGs is of potentially
great interest for SEM users as well. In this vignette, I am going to show some
possibilities. This follows the structure of Kyono's "Commentator" program, and thereby
also shows how the tasks implemented in that program can be solved using the dagitty
package.

```{r}
g1 <- dagitty( "dag {
	W1 -> Z1 -> X -> Y
	Z1 <- V -> Z2
	W2 -> Z2 -> Y
	X <-> W1 <-> W2 <-> Y
}")

g2 <- dagitty( "dag {
	Y <- X <- Z1 <- V -> Z2 -> Y
	Z1 <- W1 <-> W2 -> Z2
	X <- W1 -> Y
	X <- W2 -> Y
}")

plot(graphLayout(g1))
```

## List testable implications of a structural equation model
```{r} 
print( impliedConditionalIndependencies( g1 ) )
```

## List adjustment sets for specific path coefficients 

```{r}
print( adjustmentSets( g1, "Z1", "X", effect="direct" ) )
```

```{r}
print( adjustmentSets( g2, "X", "Y", effect="direct" ) )
```

## List path coefficients that are identifiable by regression
```{r}
for( n in names(g1) ){
	for( m in children(g1,n) ){
		a <- adjustmentSets( g1, n, m, "direct" )
		if( length(a) > 0 ){
			cat("The coefficient on ",n,"->",m,
				" is identifiable controlling for:\n",sep="")
			print( a, prefix=" * " )
		}
	}
}
```

## List adjustment sets for specific total effects

```{r}
print( adjustmentSets( g1, "X", "Y" ) )
```

```{r}
print( adjustmentSets( g2, "X", "Y" ) )
```

## List total effects that are identifiable by regression
```{r}
for( n in names(g1) ){
	for( m in setdiff( descendants( g1, n ), n ) ){
		a <- adjustmentSets( g1, n, m )
		if( length(a) > 0 ){
			cat("The total effect of ",n," on ",m,
				" is identifiable controlling for:\n",sep="")
			print( a, prefix=" * " )
		}
	}
}
```

## List path coefficients that are identifiable through instrumental variables
```{r}
for( n in names(g1) ){
	for( m in children(g1,n) ){
		iv <- instrumentalVariables( g1, n, m )
		if( length( iv ) > 0 ){
			cat( n, m, "\n" )
			print( iv , prefix=" * " )
		}
	}
}
